\Huge State-Space Models and Particle Filters

### 2021-07-30 prova_knit_finale.tex
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

@@

@@

@@

@@

In this section we present a practical illustration of the algorithms
discussed. For the sake of simplicity we use a random walk plus noise
model, i.e.~the most basic form of a linear Gaussian state-space
model.\\
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align} As already mentioned before, in this case the filtering
distribution can be computed in closed form solutions using the Kalman
filter. However, this toy example will be used also to illustrate more
involved filtering strategies described in this work. We believe indeed
that it represents a useful starting point to understand the logic of
the algorithms which may be eventually replicated when dealing with more
complex models.\\
The filtering strategy is applied to 50 simulated data. Figure XX shows
the simulated true states sequence assuming as data generating process
the Equation (2) with \(\tau^{2}=1\) and simulated observed sequence
process form Equation (1) with \(\sigma^{2}=1\).

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-5-1}

@@

\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-5}
\end{figure}

@@

\texttt{data} ~~the observed process. It has to be a vector or a
univariate time series.\\
\texttt{sig2} ~~the variance \(\sigma^{2}\) in Equation (1)\\
\texttt{tau2} ~~the variance \(\tau^{2}\) in Equation (2)\\
\texttt{m0} ~~central value of the normal prior state distribution\\
\texttt{C0} ~~variance of the normal prior state distribution

@@

In Figure XX below filtered states estimated using Kalman Filter with
\(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are compared to
the true states values. Notice how closely the filtered states follow
the observations and the goodness of the approximation of the true
states. 95 percent credible intervals are computed as
\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-7-1}

@@

\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-7}
\end{figure}

The discussion of Kalman Filter will continue in Section XX where we
compare it to the Particle Filter.

@@

Consider again the random walk plus noise model of section XX,the
challenge here is to estimate the filtered states using a Sequential
Importance Sampling algorithm. The main idea of the SIS applied to this
univariate linear gaussian model is described in the following
algorithm. We indicate with \(n\) the sample size of time observations
and with \(N\) the generated sample size for each step of the Sequenial
Monte Carlo.

@@

\texttt{data} ~~the observed process. It has to be a vector or a
univariate time series.\\
\texttt{N} ~~number of particles generated at each step\\
\texttt{m0} ~~central value of the normal prior state distribution\\
\texttt{C0} ~~variance of the normal prior state distribution\\
\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{SISfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma)\{}
\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
\NormalTok{    x    }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x,tau)                   }\CommentTok{\#sample from N(x\_\{t{-}1\},tau)}
\NormalTok{    w    }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)         }\CommentTok{\#update weight}
\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}

\NormalTok{    wnorm}\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)                         }\CommentTok{\#normalized weight}
\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(wnorm}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)                   }\CommentTok{\#effective sample size}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-10-1}

@@

\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-10}
\end{figure}

@@

In this section, Bootstrap Particle Filter will be used to estimate
filtered states of the Random Walk plus Noise introduced in section XX.
The overall strategy replicate the SIS filter with the addition of a
ESS-based resampling step that applies when the effective sample size is
smaller than a predetermined threshold opportunely chosen (in our
example we decide to follow a common rule of thumb consisting in setting
the threshold at \(N/2\)). The steps are presented in the following
algorithm.

@@

\texttt{data} ~~the observed process. It has to be a vector or a
univariate time series.\\
\texttt{N} ~~number of particles generated at each step\\
\texttt{m0} ~~central value of the normal prior state distribution\\
\texttt{C0} ~~variance of the normal prior state distribution\\
\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
otherwise, if missing, the threshold is set equal to \(N/2\)

@@

The estimated states of the Boostrap Particle Filter togheter with the
effective sample size are shown in Figure XX. We decide to set
\(N=1000\),\(m_{0}=0\),\(C_{0}=100\) and \(\tau=\sigma=1\). Notice how
the resampling step allows the effective sample size not to drop,
improving results.

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-13-1}

@@

\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-13}
\end{figure}

Since the Random Walk plus Noise model allows for closed form solutions,
we want to compare the results of Boostrap Particle Filter(BPF) with
Kalman Filter(KF). As we can see from figure XX, when the number of
particles generated at any interaction increases the results for both
the estimated mean and the variance tend to converge {[}{[}SPIEGARE
PERCHE'{]}{]}. Moreover, comparing the Root Mean Square Errors of the
filtered states with respect to the true state values, when the sample
size increases the BPF decreases and, for large N, it reaches the
accuracy of the Kalman Filter.

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-15}Root Mean Square Errors}\\
\toprule
N & Threshold & KF & BPF\\
\midrule
100 & 0.5 & 0.879 & 0.916\\
1000 & 0.5 & 0.879 & 0.882\\
10000 & 0.5 & 0.879 & 0.885\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-16-1}

@@

\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:unnamed-chunk-16}
\end{figure}

@@

\hfill\break
The Bootstrap Particle Filter Approach for the Random Walk plus Noise
Model described in Section XX can be improved accounting for the
observations in the importance transition density and it consists of
generating \(x_{t}\) from its conditional distribution given \(x_{t-1}\)
and \(y_{t}\). In the Normal model we are considering, the optimal
proposal will be a Normal density as well with mean and variance given
by \begin{align*}
\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
\end{align*} On the other hand, the incremental weights, using this
importance transition density, are proportional to the conditional
density of \(y_{t}\) given \(x_{t-1}=x_{t-1}^{(i)}\) , i.e
\(N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})\), evaluated at \(y_{t}\). In
other words the algorithm implemented in R is:

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-19-1}

@@

\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-19}
\end{figure}

Let's provide directly a brief comparison between the Bootstrap Particle
Filter (BPF) and the Guided Particle Filter (GPF). As we can see from
the Figure XX, the Guided Particle Filter provides point estimates that
are slightly better with respect to the ones of the BPF, and this is
confirmed by the Table showing the RMSE. Moreover, also the variance is
better, suggesting for higher precision.

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-21}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & GPF\\
\midrule
1000 & 0.50 & 0.882 & 0.874\\
1000 & 0.25 & 0.889 & 0.874\\
1000 & 0.10 & 0.885 & 0.879\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-22-1}

@@

\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-22}
\end{figure}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-25-1}

@@

\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-25}
\end{figure}

Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap
Particle Filter (BPF).

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-26-1}

@@

\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-26}
\end{figure}

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-28}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & APF\\
\midrule
1000 & 0.50 & 0.882 & 0.899\\
1000 & 0.25 & 0.889 & 0.883\\
1000 & 0.10 & 0.885 & 0.877\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-32-1}

@@

\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-32}
\end{figure}

@@

\hfill\break
Consider the linear Gaussian example of section XX, but this time with
unknown variances \(\tau^{2}\) and \(\sigma^{2}\). Thus, let
\(\psi=(\sigma^{2},\tau^{2})\) be the unknown parameter vector and
assign a gamma prior for its components, \begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*} Alternatively, assign them a uniform prior if we have no
knowledge of the hyperparameters. The algorithm follows the following
steps.

@@

{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-35-1}

@@

\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-35}
\end{figure}

@@

\begin{longtable}[t]{cccccc}
\caption{\label{tab:unnamed-chunk-37}RMSE}\\
\toprule
N & Threshold & KF & BPF & APF & LWF\\
\midrule
100 & 0.50 & 0.879 & 0.916 & 0.881 & 0.883\\
1000 & 0.50 & 0.879 & 0.882 & 0.899 & 0.977\\
10000 & 0.50 & 0.879 & 0.885 & 0.874 & 0.945\\
1000 & 0.50 & 0.879 & 1.172 & 1.677 & 1.140\\
1000 & 0.25 & 0.879 & 1.085 & 0.974 & 3.419\\
\addlinespace
1000 & 0.10 & 0.879 & 4.285 & 0.994 & 1.142\\
\bottomrule
\end{longtable}

@@

\chapter{Applications to Stochastic Volatility Models}
\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}

>>>>

\usepackage{babel}

@@

\usepackage{amsfonts}

@@

\usepackage{mathtools}

@@

\usepackage[toc,page]{appendix}

@@

\subsection{Implementation}\label{implementation}

The aim of this subsection is to provide a practical illustration of the
algorithms discussed in the main chapters. For the sake of simplicity we
use for implementations a random walk plus noise model, i.e.~the most
basic form of linear Gaussian state-space model.\\
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \label{eq:1} \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \label{eq:2} \\
x_{0} & \sim N(m_{0},C_{0})
\end{align} As already mentioned before, in this case the filtering
distribution can be computed in closed form solutions using the Kalman
filter. However, this toy example will be used also to illustrate more
involved filtering strategies described in this work. We believe indeed
that it represents a useful starting point to understand the logic of
the algorithms which may be eventually replicated when dealing with more
complex models.\\
The filtering strategy is applied to 50 simulated data. Figure
\ref{fig:myfig} shows the simulated true states sequence assuming as
data generating process the Equation \eqref{eq:2} with \(\tau^{2}=1\)
and simulated observed sequence process form Equation \eqref{eq:1} with
\(\sigma^{2}=1\).

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig-1}

@@

\caption{ Simulated random walk plus noise model}\label{fig:myfig}
\end{figure}

@@

\texttt{data} ~~the observed process. It has to be a vector or a
univariate time series.\\
\texttt{sig2} ~~the variance \(\sigma^{2}\) in observation equation\\
\texttt{tau2} ~~the variance \(\tau^{2}\) in state equation\\
\texttt{m0} ~~central value of the normal prior state distribution\\
\texttt{C0} ~~variance of the normal prior state distribution

@@

In Figure \ref{fig:myfig2} below filtered states estimated using Kalman
Filter with \(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are
compared to the true states values. Notice how closely the filtered
states follow the observations and the goodness of the approximation of
the true states. 95 percent credible intervals are computed as
\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig2-1}

@@

\caption{Kalman Filtered States with credible interval (in red)}\label{fig:myfig2}
\end{figure}

The discussion of Kalman Filter will continue in the next sections where
we compare it to the Particle Filter.

@@

Consider again the random walk plus noise model of section
\ref{implementation},the challenge here is to estimate the filtered
states using a Sequential Importance Sampling algorithm. The main idea
of the SIS applied to this univariate linear gaussian model is described
in the following algorithm. We indicate with \(n\) the sample size of
time observations and with \(N\) the generated sample size for each step
of the Sequenial Monte Carlo.

@@

\texttt{data} ~~the observed process. It has to be a vector or a
univariate time series.\\
\texttt{N} ~~number of particles generated at each step\\
\texttt{m0} ~~central value of the normal prior state distribution\\
\texttt{C0} ~~variance of the normal prior state distribution\\
\texttt{tau} ~~the standard deviation \(\tau\) in state equation\\
\texttt{sigma} ~~the standard deviation \(\sigma\) in observation
equation

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{SISfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma)\{}
\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
\NormalTok{    x    }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x,tau)                   }
\NormalTok{    w    }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)         }
\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}

\NormalTok{    wnorm}\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)                         }
\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(wnorm}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)                   }

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-8-1}

@@

\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-8}
\end{figure}

@@

In this section, Bootstrap Particle Filter will be used to estimate
filtered states of the Random Walk plus Noise introduced in section
\ref{implementation}. The overall strategy replicate the SIS filter with
the addition of a ESS-based resampling step that applies when the
effective sample size is smaller than a predetermined threshold
opportunely chosen (in our example we decide to follow a common rule of
thumb consisting in setting the threshold at \(N/2\)). The steps are
presented in the following algorithm.

@@

\texttt{data} ~~the observed process. It has to be a vector or a
univariate time series.\\
\texttt{N} ~~number of particles generated at each step\\
\texttt{m0} ~~central value of the normal prior state distribution\\
\texttt{C0} ~~variance of the normal prior state distribution\\
\texttt{tau} ~~the standard deviation \(\tau\) in state equation\\
\texttt{sigma} ~~the standard deviation \(\sigma\) in observation
equation\\
\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
otherwise, if missing, the threshold is set equal to \(N/2\)

@@

The estimated states of the Boostrap Particle Filter togheter with the
effective sample size are shown in Figure \ref{fig:myfig3}. We decide to
set \(N=1000\),\(m_{0}=0\),\(C_{0}=100\) and \(\tau=\sigma=1\). Notice
how the resampling step allows the effective sample size not to drop,
improving results.

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig3-1}

@@

\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:myfig3}
\end{figure}

Since the Random Walk plus Noise model allows for closed form solutions,
we want to compare the results of Boostrap Particle Filter(BPF) with
Kalman Filter(KF). As we can see from figure \ref{fig:myfig4}, when the
number of particles generated at any interaction increases the results
for both the estimated mean and the variance tend to converge
{[}{[}SPIEGARE PERCHE'{]}{]}. Moreover, comparing the Root Mean Square
Errors of the filtered states with respect to the true state values,
when the sample size increases the BPF decreases and, for large N, it
reaches the accuracy of the Kalman Filter.

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-12}Root Mean Square Errors}\\
\toprule
N & Threshold & KF & BPF\\
\midrule
100 & 0.5 & 0.879 & 0.916\\
1000 & 0.5 & 0.879 & 0.882\\
10000 & 0.5 & 0.879 & 0.885\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig4-1}

@@

\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:myfig4}
\end{figure}

@@

\hfill\break
The Bootstrap Particle Filter Approach for the Random Walk plus Noise
Model described in Section \ref{implementation} can be improved
accounting for the observations in the importance transition density and
it consists of generating \(x_{t}\) from its conditional distribution
given \(x_{t-1}\) and \(y_{t}\). In the Normal model we are considering,
the optimal proposal will be a Normal density as well with mean and
variance given by \begin{align*}
\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
\end{align*} On the other hand, the incremental weights, using this
importance transition density, are proportional to the conditional
density of \(y_{t}\) given \(x_{t-1}=x_{t-1}^{(i)}\) , i.e
\(N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})\), evaluated at \(y_{t}\). In
other words the algorithm implemented in R is:

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-15-1}

@@

\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-15}
\end{figure}

Let's provide directly a brief comparison between the Bootstrap Particle
Filter (BPF) and the Guided Particle Filter (GPF). As we can see from
the Figure \ref{fig:myfig5}, the Guided Particle Filter provides point
estimates that are slightly better with respect to the ones of the BPF,
and this is confirmed by the Table showing the RMSE. Moreover, also the
variance is better, suggesting for higher precision.

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-17}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & GPF\\
\midrule
1000 & 0.50 & 0.882 & 0.874\\
1000 & 0.25 & 0.889 & 0.874\\
1000 & 0.10 & 0.885 & 0.879\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig5-1}

@@

\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:myfig5}
\end{figure}

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-20-1}

@@

\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-20}
\end{figure}

Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap
Particle Filter (BPF). From Figure \ref{fig:myfig6} it is difficult to
say which of the two is better and the tables does not help much since
the APF apparently outperforms the BPF in two cases over three. For this
reason we can conclude that for this model the Auxiliary Particle
Filtering strategy does not provide any serious improvement to a the
basic one.

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig6-1}

@@

\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:myfig6}
\end{figure}

\begin{longtable}[t]{cccc}
\caption{\label{tab:unnamed-chunk-22}Root Mean Square Errors}\\
\toprule
N & Threshold & BPF & APF\\
\midrule
1000 & 0.50 & 0.882 & 0.899\\
1000 & 0.25 & 0.889 & 0.883\\
1000 & 0.10 & 0.885 & 0.877\\
\bottomrule
\end{longtable}

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-26-1}

@@

\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-26}
\end{figure}

@@

\hfill\break
Consider the linear Gaussian example of section \ref{implementation},
but this time with unknown variances \(\tau^{2}\) and \(\sigma^{2}\).
Thus, let \(\psi=(\sigma^{2},\tau^{2})\) be the unknown parameter vector
and assign a gamma prior for its components, \begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*} Alternatively, assign them a uniform prior if we have no
knowledge of the hyperparameters. The algorithm follows the following
steps.

@@

{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-29-1}

@@

\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-29}
\end{figure}

@@

\begin{longtable}[t]{cccccc}
\caption{\label{tab:unnamed-chunk-31}RMSE}\\
\toprule
N & Threshold & KF & BPF & APF & LWF\\
\midrule
100 & 0.50 & 0.879 & 0.916 & 0.881 & 0.883\\
1000 & 0.50 & 0.879 & 0.882 & 0.899 & 0.977\\
10000 & 0.50 & 0.879 & 0.885 & 0.874 & 0.945\\
1000 & 0.50 & 0.879 & 1.172 & 1.677 & 1.140\\
1000 & 0.25 & 0.879 & 1.085 & 0.974 & 3.419\\
\addlinespace
1000 & 0.10 & 0.879 & 4.285 & 0.994 & 1.142\\
\bottomrule
\end{longtable}

@@

\section{Implementations for Stochastic Volatility Models}

\chapter{Applications to Stochastic Volatility Models}

\section{Overview}

In the following sections, we will try to replicate some of the analyses
which were previously conducted on synthetic data, generated by a basic
stochastic volatility model; this time, the same model specification and
filters will be used on real data. However, as we will see, real data
require different strategies to evaluate the effectiveness of several
Sequential Monte Carlo techniques. Intuitively, such differences are due
to the fact that now the latent process driving the Hidden Markov Model
is not observed, nor do we know exactly its law of motion.\\
This part of the analysis is structured as follows: section
\textit{fill with the number of the section} provides a review of recent
research in the stochastic volatility literature, including extensions
of the basic specification and the use of particle filters; section
\textit{fill with the number of the section} describes the data which we
employ and the transformations that were applied to them. In section
\textit{fill with the number of the section} we briefly compare the
series of filtered states that are obtained from different particle
filters; moreover, several PFs are compared in terms of computational
cost in approximating the exact filtering distribution implied by our SV
model. Section \textit{fill with the number of the section} features a
forecast comparison between two SV specifications and a model that
assumes constant mean and variance for the financial returns, namely a
forecasting exercise that would not be possible were it not for particle
filtering techniques. In section
\textit{fill with the number of the section} we conclude by employing a
non-parametric measure of realised volatility as a proxy for the latent
volatility process, allowing a RMSE and MAE comparison between filters
which is similar to that of section
\textit{fill with the number of the section} for synthetic data.

\section{A review of Stochastic Volatility models}

In section \textit{fill with the number of the section} a simple
specification of a Stochastic Volatility model was presented and used as
the data generating process of a simulated dataset. Denoting by \(y_t\)
the log returns at time \(t\) and by \(x_t\) the latent volatility
stochastic process, such state-space model was presented as follows,
\begin{align*}
    y_t|x_t & \sim N(0,e^{x_t})\\
    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align*}

This specification - which we will employ in this section on a real
dataset - can be deemed as the standard version of the SV model.
Throughout the recent decades, several extensions have been proposed,
mostly (but not exclusively) in the field of financial econometrics. On
the one hand, for simplicity we will stick to the standard version, one
that will suffice for the scope of this work on Sequential Monte Carlo
methods. On the other hand, it is worth mentioning some of these
extensions: indeed, one can rather safely assume that, through their
higher degree of sophistication, such extensions might well improve on
the performance of our standard SV model in the analyses that will
follow.\\
~\\
Stochastic Volatility Models, whose early formulation is commonly
attributed to Taylor (1982, 1986), allow to account for time-varying and
autoregressive volatility in financial returns, posing themselves as a
valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models
in dealing with non-constant volatility. Kim, Shephard and Chib (1998)
define the canonical model for regularly spaced data as \begin{align*}
    y_t & = \psi e^{\frac{x_t}{2}}\epsilon_t\\
    x_{t+1} & = \mu+\beta(x_t-\mu)+\tau \eta_t\\
    x_1|\sigma,\beta & \sim  N\Big(\mu,\frac{\sigma^2}{1-\beta^2}\Big)\\
    \epsilon_t & \sim N(0,1)\\ 
    \eta_t & \sim N(0,1)
\end{align*} where the log volatility \(x_t\) is assumed to follow a
stationary process (\(|\beta|<1\)), \(\psi\) is a constant scaling
factor, \(\beta\) is the persistence in the volatility and \(\tau\) is
the volatility of the
log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$. Kim, Shephard and Chib (1998) prefer $\psi=1$.}
Most importantly, \(\epsilon_t\) and \(\eta_t\), the Gaussian white
noise processes that drive the canonical model, are assumed to be
uncorrelated. This latter assumption justifies another definition of
this specification, namely ``discrete SV model without
leverage''.\footnote{Note that in this review we do not focus on continuous time SV models. In fact, such models have attracted a considerable amount of research in financial econometrics and mathematical finance, especially after Hull and White (1987) considered stochastic volatility for option pricing. Arguably, the most influential model was then proposed for option pricing by Heston (1993), a SV model with leverage effects and square root diffusion driving volatility. Diffusion-based SV models enjoyed increasing popularity, see for example Barndorff-Nielsen and Shephard (2001), or Christoffersen et al. (2010), who investigate alternatives to the entrenched affine square root SV model.  Eraker (2004) proposed a SV model with correlated jumps in prices and volatility, extending Heston's model, while Comte and Renault (1998) extended Hull and White's model as to feature long memory properties. More recently, Gatheral et al. (2018) further built on Comte and Renault's fractional SV model to propose the popular "rough volatility models" (see also Friz et al., 2021).}\\
In order to accommodate for leverage effects, namely capture the
increase in volatility that follows a drop in the returns, the model can
extended as in Omori et al.~(2007), \begin{align*}
    y_t & = e^{\frac{x_t}{2}}\epsilon_t\\
    x_{t+1} & = \mu+\beta(x_t-\mu)+\eta_t\\
    \begin{pmatrix}
       \epsilon_t \\
       \eta_t
    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\Sigma), \ \Sigma=
    \begin{pmatrix}
       1 & \rho\tau \\
       \rho\tau & \tau^2
    \end{pmatrix}
\end{align*} where \(\rho<0\) captures the negative correlation. Such
specification, which can be referred to as ``discrete SV model with
leverage'', captures the asymmetric response of volatility to returns of
different signs, so that similar specifications are sometimes also
deemed as ``asymmetric SV models'' (e.g.~Harvey and Shephard 1996, Mao
et al.~2020).\\
Note that so far the assumptions on \(\epsilon_t\) and \(\eta_t\)
implied that the returns are conditionally normally distributed. The
discrete time SV models can also be extended to allow for heavy-tailed
or asymmetric conditional returns distributions: symmetric or skewed
Student-\(t\), Generalised Hyperbolic (GH) distribution, Generalised
Error Distribution (GED) and scale mixtures of normals feature as
popular choices (Kim and Stoffer 2008, Nakajima and Omori 2012, Mao et
al.~2020). In fact, SV with heavy tailed return distributions were shown
to better meet empirical regularities like the leptokurtic distribution
of the returns and slowly decaying autocorrelation functions of the
squared returns (Liesenfeld and Jung 2000).\\
Assuming that \(\epsilon_t\) follows a Student-\(t\) distribution, and
exploiting the fact that \(\epsilon_t\) can then be written as
\(\lambda_t^{-1/2}\zeta_t\), where \(\zeta_t \sim N(0,1)\) and
\(v\lambda_t \sim \chi^2_v\) (Harvey et al.~1994, Chib et al.~2002), we
have the following SV model with both fat tails and leverage effect
(Jacquier et al., 2004), \begin{align*}
    y_t & = e^{\frac{x_t}{2}}\lambda_t^{-1/2}\zeta_t\\
    x_{t+1} & = \alpha + \beta x_t + \eta_t\\
    \begin{pmatrix}
       \zeta_t \\
       \eta_t
    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\Sigma), \ \Sigma=
    \begin{pmatrix}
       1 & \rho\tau \\
       \rho\tau & \tau^2
    \end{pmatrix}\\
    v\lambda_t & \sim \chi^2_v
\end{align*} To capture other elements of the behaviour of financial
data, several other extensions have been proposed. For example, SV
models have been extended to include conditional heteroskedasticity in
the mean returns (Koopman and Hol Uspensky 2002) to capture potential
volatility feedback effects, or to feature autoregressive moving average
innovations (Chan 2013, Zhang et al.~2020), allowing better goodness of
fit and out-of-sample
